Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-07-06 manuscript/reviews/ResponseReview1.Rmd

Response: Thanks for your constructive and encouraging feedback. As we also noted in our response to the editor, we have made substantial revisions to the manuscript that include removing repetitive text, and placing a stronger focus on clarity and brevity in the revised ms.

> Major comments:
1. Some discussion on the choice of bin width is needed. Since all of the example
time-to-event data is collected in continuous time, the analyst must discretize it
by defining bins. Defining bins is a key step in the analysis pipeline and so
should not be overlooked. It seems like results—and potentially conclusions
drawn—may depend on the choice of bin width. What factors should the analyst
consider when defining the width of the bins? Does the number of bins matter?
Or is the number of events per bin important? What about other factors? Should
analysts (or potentially these authors too) consider multiple bin widths when
conducting an analysis? An example comparing results when bin width varies
could be both interesting and informative.

Response: We discuss the issue of defining bins on page 12 of the revised ms.:

"Third, the width of each time bin will need to be determined. For instance, in Figure
1B we chose 100 ms in an arbitrary manner. In reality, however, bin width will need to be
set by considering a number of factors simultaneously. The optimal bin width will depend
on (a) the length of the observation period in each trial, (b) the rarity of event occurrence,
(c) the number of repeated measures (or trials) per condition per participant, and (d) the
shape of the hazard function. Finding an appropriate bin width in a given user case before
fitting models will require testing a number of options, when calculating and plotting the
descriptive statistics (see section 3.1). The goal is to find the smallest bin width that is
supported by the amount of data available. Based on our experience, a bin width of 50 ms
is a good starting value when the number of repeated measures is 100 or less. Too small
bin widths will result in erratic hazard functions as many bins will have no events, and
thus hazard estimates of zero. Interestingly, the time bins do not need to have the same
width. For example, Panis (2020) used larger bins towards the end of the observation
period, as fewer events occurred there."

@@

Response: Although SAT is rather standard in cognitive experimental psychology, we tried to limit the use of abbreviations in main body of the revised ms.

> It would be helpful if the research questions that can be addressed by the methods presented are clearly presented in each section. I think that currently they are getting lost in the tutorial.

Response: We now discuss general research questions on page 10 of the revised ms.

We did not add more information about possible research questions in each section of Tutorial 2a, because we feel that (a) research questions are highly dependent on the field of study (perception, behavioral control, memory), (b) this would make the tutorial even longer, and (c) we mention in the paper that we do not discuss **why** you might perform EHA. In other words, we think that the knowledge of experimental psychologists on regression modeling should be large enough for them to link the model structure to possible research questions, once they have read the paper and worked through the code examples.

### 2025-07-06 manuscript/reviews/ResponseReview1.tex

Response: Thanks for your constructive and encouraging feedback. As we
also noted in our response to the editor, we have made substantial
revisions to the manuscript that include removing repetitive text, and
placing a stronger focus on clarity and brevity in the revised ms.

\begin{quote}
Major comments: 1. Some discussion on the choice of bin width is needed.
Since all of the example time-to-event data is collected in continuous
time, the analyst must discretize it by defining bins. Defining bins is
a key step in the analysis pipeline and so should not be overlooked. It
seems like results---and potentially conclusions drawn---may depend on
the choice of bin width. What factors should the analyst consider when
defining the width of the bins? Does the number of bins matter? Or is
the number of events per bin important? What about other factors? Should
analysts (or potentially these authors too) consider multiple bin widths
when conducting an analysis? An example comparing results when bin width
varies could be both interesting and informative.
\end{quote}

Response: We discuss the issue of defining bins on page 12 of the
revised ms.:

``Third, the width of each time bin will need to be determined. For
instance, in Figure 1B we chose 100 ms in an arbitrary manner. In
reality, however, bin width will need to be set by considering a number
of factors simultaneously. The optimal bin width will depend on (a) the
length of the observation period in each trial, (b) the rarity of event
occurrence, (c) the number of repeated measures (or trials) per
condition per participant, and (d) the shape of the hazard function.
Finding an appropriate bin width in a given user case before fitting
models will require testing a number of options, when calculating and
plotting the descriptive statistics (see section 3.1). The goal is to
find the smallest bin width that is supported by the amount of data
available. Based on our experience, a bin width of 50 ms is a good
starting value when the number of repeated measures is 100 or less. Too
small bin widths will result in erratic hazard functions as many bins
will have no events, and thus hazard estimates of zero. Interestingly,
the time bins do not need to have the same width. For example, Panis
(2020) used larger bins towards the end of the observation period, as
fewer events occurred there.''

@@

Response: Although SAT is rather standard in cognitive experimental
psychology, we tried to limit the use of abbreviations in main body of
the revised ms.

\begin{quote}
It would be helpful if the research questions that can be addressed by
the methods presented are clearly presented in each section. I think
that currently they are getting lost in the tutorial.
\end{quote}

Response: We now discuss general research questions on page 10 of the
revised ms.

We did not add more information about possible research questions in
each section of Tutorial 2a, because we feel that (a) research questions
are highly dependent on the field of study (perception, behavioral
control, memory), (b) this would make the tutorial even longer, and (c)
we mention in the paper that we do not discuss \textbf{why} you might
perform EHA. In other words, we think that the knowledge of experimental
psychologists on regression modeling should be large enough for them to
link the model structure to possible research questions, once they have
read the paper and worked through the code examples.

### 2025-07-06 manuscript/reviews/ResponseReview1.Rmd

> 5. Section 2.3.2. I like the idea of making a connection with experimental design,
but am having trouble seeing how the discussion in this section is specific to EHA
and how these comments do not apply also to the “orthodox method”. I think this
section could be improved by providing more specific distinctions between
experimental design implications for EHA vs. the “orthodox method”.

Response: We created a new Figure 1 that makes the intended point, and removed section 2.3.2.

> 6. Line 517: What are model weights? Please explain briefly for the reader,
including how the interpretation differs for loo and waic. 
Also, in the provided
reference (Kruz 2023a), I was only able to find a discussion of dWAIC, rather
than WAIC, so could you please include a definition of or reference for dWAIC in
your context? 
Lastly, when reporting the model weights in the example in the
text, I’d suggest including more than 2 digits so that it is clear if the reported
weights are actually 0 and 1 or if rounding has just made them appear so.

Response: We briefly explain model weights for loo and waic on page 28 of the revised ms.
A discussion of WAIC in Kurz (2023a) can be found in section 4.6.4., so we did not include a reference for dWAIC in the revised ms. Lastly, we print more digits when reporting the model weights.

@@

> 8. Table 2: It could be helpful to show an example of a trial where an event never
occurs, to help the reader connect what is written in the text (line 325-327) to
what appears in the table.

Response: We added data visualizations in section A of the Supplemental Material.

> 9. Figure 4: Why is there no bin 40?

Response: Because no responses occurred in the first bin with endpoint 40 ms. However, we added this bin back in Figure 2 (previously 4) of the revised ms to avoid confusion.

@@

> <b>Tutorials section:</b>
-Rather than only using h(t), and ca(t), suggest to use hazard function and conditional accuracy function along with their mathematical notation. Without having presented any equations, using terms such as h(t) and ca(t) seems odd.

Response: We try to limit the number of abbreviations used in the main body of the revised ms.

> Similarly, P(t) is never defined for readers.

Response: P(t) was and still is defined only in the Supplementary Material (section B in the revised version).

@@

> section 4.3.1 (line 453) – more detail is needed on the cloglog link function and how it differs from the more commonly known about logit link.

Response: A visual comparison between the cloglog and logit link functions is provided in section C of the Supplemental Material.

> section 4.3.1 (line 456) – what goes into the decision about whether to treat time as a categorical or continuous predictor?

Response: We added the following information on page 24 of the revised ms.:
"Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor (see also section E of the Supplemental Material). For example, when you want to know if cloglog-hazard is changing linearly or quadraticly over time, you should treat TIME as a continuous predictor.
When you are only interested in the effect of covariates on hazard, you can treat TIME as a categorical predictor (i.e., fit an intercept for each bin), in which case you can choose between reference coding and index coding."

@@

> section 4.3.3. – it might be helpful to give a brief indicator up front to readers about the rationale/purpose of fitting both M0i and M1i, especially given that no interpretation is given for M0i. Similarly, it would be helpful to give a bit more detail on the purpose of the comparison of M0i and M1i (given that this is a didactic tutorial).

Response: We mention now that model M0i is a baseline or reference model, to be compared with model M01 (see page 26 of the revised ms.). We also mention that we only report 2 models in the paper from a larger set of models that are fitted in the file Tutorial_2a.Rmd.

@@

> pg. 36 – it would be helpful to give more detail on how to interpret the logit-ca scale

Response: We decided not to include more detail because logistic regression is very familiar to experimental psychologists.

> section 4.7 – as previously mentioned, it is possible that tutorial 4 could be a stand alone tutorial. If it’s not though, more details are needed up front to tell readers what the purpose is and why this is useful.

Response: We decided to keep tutorial 4, and discuss its usefulness on pages 45-46 of the revised ms.

### 2025-07-06 manuscript/reviews/ResponseReview1.tex

\begin{quote}
section 4.3.1 (line 453) -- more detail is needed on the cloglog link
function and how it differs from the more commonly known about logit
link.
\end{quote}

Response: A visual comparison between the cloglog and logit link
functions is provided in section C of the Supplemental Material.

\begin{quote}
section 4.3.1 (line 456) -- what goes into the decision about whether to
treat time as a categorical or continuous predictor?
\end{quote}

Response: We added the following information on page 24 of the revised
ms.: ``Third, one has to choose whether to treat TIME (i.e., the time
bin index t) as a categorical or continuous predictor (see also section
E of the Supplemental Material). For example, when you want to know if
cloglog-hazard is changing linearly or quadraticly over time, you should
treat TIME as a continuous predictor. When you are only interested in
the effect of covariates on hazard, you can treat TIME as a categorical
predictor (i.e., fit an intercept for each bin), in which case you can
choose between reference coding and index coding.''

@@

\begin{quote}
section 4.3.3. -- it might be helpful to give a brief indicator up front
to readers about the rationale/purpose of fitting both M0i and M1i,
especially given that no interpretation is given for M0i. Similarly, it
would be helpful to give a bit more detail on the purpose of the
comparison of M0i and M1i (given that this is a didactic tutorial).
\end{quote}

Response: We mention now that model M0i is a baseline or reference
model, to be compared with model M01 (see page 26 of the revised ms.).
We also mention that we only report 2 models in the paper from a larger
set of models that are fitted in the file Tutorial\_2a.Rmd.

### 2025-07-06 manuscript/manuscript.Rmd
update ms. and Response to reviewers

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, because the first few bins typically contain no responses, one has to select an analysis time window, i.e., a contiguous set of bins for which there is data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see section D of the Supplemental Material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data [@singerAppliedLongitudinalData2003]. Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor (see also section E of the Supplemental Material). For example, when you want to know if cloglog-hazard is changing linearly or quadracticly over time, you should treat TIME as a continuous predictor.
When you are only interested in the effect of covariates on hazard, you can treat TIME as a categorical predictor (i.e., fit an intercept for each bin), in which case you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial_2a.Rmd for an example). With index coding, one constructs an index variable that contains integers that correspond to different categories (see models M0i and M1i below). As explained by @mcelreathStatisticalRethinkingBayesian2020, the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.

@@

The middle column of Supplementary Figure 2 (section D of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 2 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

>>>>

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, because the first few bins typically contain no responses, one has to select an analysis time window, i.e., a contiguous set of bins for which there is data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see section D of the Supplemental Material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data [@singerAppliedLongitudinalData2003]. Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor (see also section E of the Supplemental Material). For example, when you want to know if cloglog-hazard is changing linearly or quadraticly over time, you should treat TIME as a continuous predictor.
When you are only interested in the effect of covariates on hazard, you can treat TIME as a categorical predictor (i.e., fit an intercept for each bin), in which case you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial_2a.Rmd for an example). With index coding, one constructs an index variable that contains integers that correspond to different categories (see models M0i and M1i below). As explained by @mcelreathStatisticalRethinkingBayesian2020, the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.

@@

The middle column of Supplementary Figure 3 (section F of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 3 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

### 2025-07-06 manuscript/manuscript.tex
update ms. and Response to reviewers

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.5.1; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.13.0; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.9.0.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.2; Wickham, 2016), \emph{lme4} (Version 1.1.37; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.4; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.3; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.168; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.4; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.14; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{rstan} (Version 2.32.7; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.3.0; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.5; Barth, 2023) for all reported analyses.

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, because the first few bins typically contain no responses, one has to select an analysis time window, i.e., a contiguous set of bins for which there is data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see section D of the Supplemental Material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor (see also section E of the Supplemental Material). For example, when you want to know if cloglog-hazard is changing linearly or quadracticly over time, you should treat TIME as a continuous predictor.
When you are only interested in the effect of covariates on hazard, you can treat TIME as a categorical predictor (i.e., fit an intercept for each bin), in which case you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial\_2a.Rmd for an example). With index coding, one constructs an index variable that contains integers that correspond to different categories (see models M0i and M1i below). As explained by McElreath (2020), the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.

>>>>

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.5.1; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.13.0; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.9.0.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.2; Wickham, 2016), \emph{lme4} (Version 1.1.37; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.4; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.3; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.168; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.4; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.14; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{rstan} (Version 2.32.7; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.3.0; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.5; Barth, 2023) for all reported analyses.

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, because the first few bins typically contain no responses, one has to select an analysis time window, i.e., a contiguous set of bins for which there is data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see section D of the Supplemental Material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor (see also section E of the Supplemental Material). For example, when you want to know if cloglog-hazard is changing linearly or quadraticly over time, you should treat TIME as a continuous predictor.
When you are only interested in the effect of covariates on hazard, you can treat TIME as a categorical predictor (i.e., fit an intercept for each bin), in which case you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial\_2a.Rmd for an example). With index coding, one constructs an index variable that contains integers that correspond to different categories (see models M0i and M1i below). As explained by McElreath (2020), the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.
